---
title: 'EDS 223: Discussion Week 6 - Mapping Human Modification onto California Protected Areas'
author: 'Name'
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Prerequsites 

```{r}

```

Load in the CPAD_2023a_SuperUnits.shp and the ghm.tif files. ghm.tif will take 5-8 minutes. Transform both to EPSG:4326.

```{r}

```

While this loads - think about what you learned about raster/vector interactions last week. Discussion in your groups how you think you would combine these two datasets. 

## Exercises

1. Plot California protected areas by access level and by layer

```{r}

```

```{r}

```

2. Plot the Global human modification layer. *Hint: just use plot for ease*

```{r}

```

3. Crop and mask the raster to fit the area of the shapefile. Plot each step. *Hint: just use plot for ease*

```{r}

```

4. Extract the global human modification values at each California protected area.

```{r}

```

5. Summarize the mean of human modification within each protected area

```{r}

```

6.  Join this summary with the protected area database. Plot these values.

```{r}

```

```{r}

```

7. Join the entire human modification layer to the protected area database. Remove geometry and any missing values.

```{r}

```

8. Create two boxplots of human modification values, 1 across access_typ and 1 across layer. View these side by side.

```{r}

```
