---
title: 'EDS 223: Discussion Week 1 - Making Maps'
author: 'Name'
date: "2023-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html).

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(spData)
library(spDataLarge)
library(tidyverse)
library(ggspatial)
library(patchwork)
```

## Exercise 1

These exercises rely on a new data object based on the **world** and **worldbank_df** datasets from the **spData* package.

```{r include=TRUE}
africa = world |> 
  filter(continent == "Africa", !is.na(iso_a2)) |> 
  left_join(worldbank_df, by = "iso_a2") |> 
  dplyr::select(name, subregion, gdpPercap, HDI, pop_growth) |> 
  st_transform("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +lon_0=25")
```

## Map 1

Create a map showing the geographic distribution of the Human Development Index (HDI) across Africa.  

```{r include=TRUE}

```

## Map 2

Update the map created for question 1 so the legend has three bins: "high" (HDI above 0.7), "medium" (HDI between 0.55 and 0.7), and "low" (HDI below 0.55).   
-   change the legend title 
-   change the class labels   
-   change the color palette   

```{r include=TRUE}

```

## Map 3

Represent Africa's subregions on the map.  
-   change the color palette 
-   change the legend title 

```{r include=TRUE}

```

-   combine this map with the map from question 2 into a single plot   

```{r}

```


## Exercise 2

Here, we will use the **zion** and **nlcd** datasets from the **spDataLarge** package.

```{r include=TRUE}
zion = st_read((system.file("vector/zion.gpkg", package = "spDataLarge")))
data(nlcd, package = "spDataLarge")
force(nlcd)
```

First, we need to convert the raster into a dataframe

```{r}

```

## Map 4

Create a land cover map of Zion National Park    
-   change the default colors to match your perception of land cover categories   
-   move the map legend outside of the map to improve readability 
-   add a scale bar and north arrow and change the position of both to improve the maps aesthetics 
-   add the park boundaries on top of the land cover map    

```{r include=TRUE}

```

## Map 5

-   add an inset of Zion's location in the context of the state of Utah  

```{r}

```

```{r}

```

```{r}

```