---
title: 'EDS 223: Discussion Week 10 - Workflows and Maps'
author: 'Name'
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE,error=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=FALSE,message=FALSE)
```


## Introduction

## Prerequsites 

```{r}
library(tidyverse)
library(patchwork)
library(sf)
library(terra)
library(stars)
library(ggspatial)
```

Load in the CPAD_2023a_SuperUnits.shp and the ghm.tif files. ghm.tif Transform both to EPSG:4326.

```{r}
cpad_super <- st_read(here::here("discussion", "data", "CPAD_2023a_SuperUnits.shp"), quiet = TRUE) %>%
  sf::st_transform("EPSG:4326") %>%
  janitor::clean_names() %>%
  mutate(ID = row_number())

ghm <- rast(here::here("discussion", "data", "gHM_masked.tif")) %>%
  project("EPSG:4326")
```

## Exercises

1. Let’s make nice plots of the California Protected areas by access level

```{r}

```

```{r}

```

2. Let's try plotting the ghm layers nicely too!

```{r}

```

3. Create a function to take 2 data sets (1 polygon and 1 raster) and create a boxplot of the values based on a specific layer

```{r}
summary_boxplot = function(polygon, raster, my_layer, my_label) {
  
  # rasterize polygon by layer

  
  #do mean zonal statistics

  #join with polygon database

  
  #create boxplot based on your layer

}
```

4. Let's select some layers and use our new function!

```{r}

```

```{r}

```

```{r}

```
