---
title: 'EDS 223: Discussion Discussion Week 9 - Functions for HW4'
author: 'Name'
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE,error=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=FALSE,message=FALSE)
```

## Introduction

## Prerequsites 

```{r}
library(tidyverse)
library(sf)
library(terra)
library(kableExtra)
```

Load in the CPAD_2023a_SuperUnits.shp and the gHM_masked.tif files.

```{r}
cpad_super <- st_read(here::here("discussion", "data", "CPAD_2023a_SuperUnits.shp"), quiet=TRUE) %>%
  sf::st_transform("EPSG:4326") %>%
  janitor::clean_names() %>%
  mutate(ID = row_number())

ghm <- rast(here::here("discussion", "data", "gHM_masked.tif")) %>%
  project("EPSG:4326")
```

```{r}

```

## Exercises

1. Let's rasterize the cpad dataset a few times! This is necessary to do zonal statistics!

```{r}

```

2. Let's say you're interested in finding protected areas with intermediate levels of human modification. How could you do this?

a) masking and zonal statistics

```{r}

```

Say you want to make this more complicated, and you want two groups?

```{r}

```

Or you just want to group them generally! 

```{r}

```

Now let's use zonal statistics to get a mean

```{r}

```

b) combining and filtering 

```{r}

```

Summarize the mean of human modification within each protected area

```{r}

```

Join this summary with the protected area database. Plot these values

```{r}

```

Now we can filter! 

```{r}

```

3. These do give slightly different results however... can you identify the difference? 

```{r}

```

```{r}

```

4. Let's use zonal statistics to summarize human modification by access type!

```{r}

```
