---
title: 'EDS 223: Discussion Week 5 - Vector Raster Interactions'
author: 'Name'
date: "2023-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from [Chapter 6 of Geocomputation with R by Rovin Lovelace](https://r.geocompx.org/raster-vector).

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(spData)
library(spDataLarge)
library(tidyverse)
```

```{r}
zion_points = read_sf(system.file("vector/zion_points.gpkg", package = "spDataLarge"))
srtm = rast(system.file("raster/srtm.tif", package = "spDataLarge"))

ch = st_combine(zion_points) |>
  st_convex_hull() |> 
  st_as_sf()

grain = rast(system.file("raster/grain.tif", package = "spData"))
```

```{r}

```

## Exercises

1. Crop the srtm raster using (1) the zion_points dataset and (2) the ch dataset. Are there any differences in the output maps? Next, mask srtm using these two datasets. Can you see any difference now?

```{r}

```

2. Subset points higher than 3100 meters in New Zealand (the nz_height object) and create a template raster with a resolution of 3 km for the extent of the new point dataset. Using these two new objects:

```{r}

```

2a. Count numbers of the highest points in each grid cell.

```{r}

```

2b. Find the maximum elevation in each grid cell.

```{r}

```

3. Aggregate the raster counting high points in New Zealand (created in the previous exercise), reduce its geographic resolution by half (so cells are 6 by 6 km) and plot the result. Resample the lower resolution raster back to the original resolution of 3 km. 

```{r}

```

4. Polygonize the grain dataset and filter all squares representing clay.

```{r}

```

