---
title: "EDS 223: week 1 lab template"
output: html_document
date: "2022-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites

```{r install, include=TRUE, echo=FALSE, eval=FALSE}
install.packages('sf')
install.packages('raster')
install.packages('dplyr')
install.packages('spData')
install.packages('tmap')
install.packages('leaflet')
install.packages('ggplot2')
```


```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
```

## tmap basics

Add fill layer to New Zealand shape
```{r include=TRUE}
tm_shape(nz) +
  tm_fill()
```
Add just borders
```{r include=TRUE}
tm_shape(nz) +
  tm_borders()
```

Add fill and border layers
```{r include=TRUE}
tm_shape(nz) +
  tm_fill() +
  tm_borders()
```

## map objects

-   **tmap** can store maps as *objects*  
-   this means that we store a base map and add additional layers later
```{r include = TRUE}
map_nz <- tm_shape(nz) +
  tm_polygons()
class(map_nz)
```

-   we can add new shapes on top of the base map
-   when we add a new shape, all subsequent aesthetic functions refer to it, until we add a new shape
-   in this case, we're adding a layer with information on elevation
-   we modify this layer to have 70% transparency
```{r include = TRUE}
map_nz1 <- map_nz +
  tm_shape(nz_elev) +
  tm_raster(alpha = 0.7)

map_nz1 
```
```{r include = TRUE}
map_nz2 <- map_nz1 +
  tm_shape(nz_height) +
  tm_dots()

map_nz2
```

