---
title: "EDS 223: week 1 lab"
author: "Ruth Oliver"
output: html_document
date: "2022-09-28"
---
## introduction
In this lab, we'll explore the basics of map-making in R using the **tmap** package.
The following materials are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## prerequisites

```{r install, include=TRUE, eval=FALSE}
install.packages('sf')
install.packages('raster')
install.packages('dplyr')
install.packages('spData')
install.packages('tmap')
install.packages('leaflet')
install.packages('ggplot2')
install.packages('gifski')
```

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
```

## map-making basics

Let's start with a pre-loaded spatial object representing the states of New Zealand

```{r include=TRUE}
nz
```

-   we're going to start by using the **tmap** package to make some basic maps
-   **tmap** can work with spatial objects of a variety of clasess, meaning it's highly versatile
-   this approach relies on a series of functions that typically start with "tm\_"

the first element is always "tm_shape"

```{r include=TRUE}
tm_shape(nz) +
  tm_fill()
```

now let's plot just the boundaries

```{r include=TRUE}
tm_shape(nz) +
  tm_borders()
```

and the shapes and boundaries together

```{r include=TRUE}
tm_shape(nz) +
  tm_fill() +
  tm_borders()
```

## map objects

-   **tmap** can store maps as *objects*
-   this means that we store a base map and add additional layers later

```{r include = TRUE}
map_nz <- tm_shape(nz) +
  tm_polygons()
class(map_nz)
```

-   we can add new shapes on top of the base map
-   when we add a new shape, all subsequent aesthetic functions refer to it, until we add a new shape

In this case, we're adding a layer with information on elevation and this layer to have 70% transparency.

```{r include = TRUE}
map_nz1 <- map_nz +
  tm_shape(nz_elev) +
  tm_raster(alpha = 0.7)

map_nz1 
```

we can add points designating high points in the country

```{r include = TRUE}
map_nz2 <- map_nz1 +
  tm_shape(nz_height) +
  tm_dots()

map_nz2
```

## aesthetic basics

-   up until now, we've been working with the default aesthetics
-   there are 2 types of aesthetics: fixed and those that change with the value of a variable
-   **tmap** works differently than **ggplot2** and doesn't use the "aes()" function

Let's start by changing some fixed aesthetics...First, let's change the color used to fill the NZ shapes.

```{r include=TRUE}
tm_shape(nz) +
  tm_fill(col = "red")
```

now change the color of the boundaries

```{r include=TRUE}
tm_shape(nz) + 
  tm_borders(col = "blue")
```

and the width of the boundary lines

```{r include=TRUE}
tm_shape(nz) + 
  tm_borders(lwd = 3)
```

and the line type of the boundary lines

```{r include=TRUE}
tm_shape(nz) + 
  tm_borders(lty = 2)
```

all together now!

```{r include=TRUE}
tm_shape(nz) + 
  tm_fill(col = "red", alpha = 0.3) +
  tm_borders(col = "blue", lwd = 3, lty = 2)
```

Now let's change the colors based on a value. We noticed that the New Zealand dataset has a column with each state's land area

```{r include=TRUE}
nz
```

Let's try to plot the Land_area column. We might think that the following works, but it doesn't!

```{r include=TRUE}
#tm_shape(nz) +
#  tm_fill(col = nz$Land_area)
```

Instead, **tmap** is expecting a character string naming the attribute associated with the geometry

```{r include = TRUE}
tm_shape(nz) +
  tm_fill(col = "Land_area")
```

We can also add an argument that updates the title of the legend

```{r include=TRUE}
tm_shape(nz) +
  tm_fill(col = "Land_area", title = "Area")
```

We can even make it more precise using the "expression" function

```{r include=TRUE}
tm_shape(nz) +
  tm_fill(col = "Land_area", title = expression("Area (km"^2*")")) +
  tm_borders()
```

## color settings

-   how we set and define colors can radically change the interpretation of our maps
-   the style argument has several options for breaking data into bins
    -   **style = "pretty"** (default) rounds breaks into evenly spaced whole numbers, where possible
    -   **style = "equal"** divides input values into bins with equal range (best for uniform distributions)
    -   **style = "quantile"** puts the same number of observations into each bin
    -   **style = "jenks"** identifies groups of similar values and maximizes differences between bins

Notice how the following maps display the same data, but look quite different    
```{r include=TRUE}
tm_shape(nz) + 
  tm_polygons(col = "Median_income", style = "pretty")
```

```{r include=TRUE}
tm_shape(nz) + 
  tm_polygons(col = "Median_income", style = "equal")
```

```{r include=TRUE}
tm_shape(nz) + 
  tm_polygons(col = "Median_income", style = "quantile")
```

```{r include=TRUE}
tm_shape(nz) + 
  tm_polygons(col = "Median_income", style = "jenks")
```

We can also define custom bins
```{r include=TRUE}
breaks = c(0, 3, 4, 5) * 10000
tm_shape(nz) + 
  tm_polygons(col = "Median_income", breaks = breaks)
```

-   in some cases we might not want to use bins
    -   **style = "cont"** displays colors as a continuous spectrum
    -   **style = "cat"** uses a unique vale for each categorical value
```{r include=TRUE}
map_nz +
  tm_shape(nz_elev) +
  tm_raster(alpha = 0.7, 
            style = "cont")
```

```{r include=TRUE}
map_nz +
  tm_shape(nz) +
  tm_polygons(col = "Island",
              style = "cat")
```   
  
## map layout
-   now that we have the basics, we can turn to all the other elements that make a cohesive map
-   **tmap** has lots of options, but we will explore just a few

To clearly give readers the context of our map, we can include a compass and scale bar
```{r include=TRUE}
map_nz +
  tm_compass(type = "4star", position = c("left", "top")) +
  tm_scale_bar(breaks = c(0, 100, 200), text.size = 1)
```

Instead of using a compass and scale bar, we could add latitude/longitudes graticules
```{r include=TRUE}
map_nz +
  tm_graticules()
```

We can also update the background color
```{r include=TRUE}
map_nz +
  tm_graticules() +
  tm_layout(bg.color = "lightblue")
```

## faceted and animated maps
-   we might have data that varies over time and we want to look at the how it changes
-   one approach is by using faceted plots, a series of plots

```{r include=TRUE}
urb_1970_2030 <- urban_agglomerations %>% 
  filter(year %in% c(1970, 1990, 2010, 2030))

tm_shape(world) +
  tm_polygons() +
  tm_shape(urb_1970_2030) +
  tm_symbols(col = "black", border.col = "white", size = "population_millions") +
  tm_facets(by = "year", nrow = 2, free.coords = FALSE)
```

-   another approach would be to create a gif that we can embed
-   in this case we use "along" instead of "by"
```{r include=TRUE, eval=FALSE}
urb_anim = tm_shape(world) + 
  tm_polygons() + 
  tm_shape(urban_agglomerations) + 
  tm_dots(size = "population_millions") +
  tm_facets(along = "year", free.coords = FALSE)

tmap_animation(urb_anim, filename = "urb_anim.gif", delay = 25)
```

## interactive maps
-   **tmap** is especially powerful because it allows us to make interactive maps using the same syntax
-   all we need to do is enter the interactive mode

```{r include=TRUE}
tmap_mode("view")
map_nz
```

To go back to regular plotting, we just need enter plotting mode
```{r include=TRUE}
tmap_mode("plot")
```

