---
title: 'EDS 223: week 8 lab'
author: "Ruth Oliver"
date: "2022-11-16"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---
# Overview
In this lab we are analyzing plant phenology in Southern California. 

credit: this lab is based on a materials developed by Chris Kibler.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=TRUE}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(here)

rm(list = ls())
here::i_am("labs/week8.Rmd")
```

#### load Landsat data
Here we load 8 scenes collected by the Landsat OLI sensor.
```{r include=TRUE}
landsat_20180612 <-rast(here("labs", "data", "week8", "landsat_20180612.tif"))
landsat_20180815 <- rast(here("labs", "data", "week8", "landsat_20180815.tif"))
landsat_20181018 <- rast(here("labs", "data", "week8", "landsat_20181018.tif"))
landsat_20181103 <- rast(here("labs", "data", "week8", "landsat_20181103.tif"))
landsat_20190122 <- rast(here("labs", "data", "week8", "landsat_20190122.tif"))
landsat_20190223 <- rast(here("labs", "data", "week8", "landsat_20190223.tif"))
landsat_20190412 <- rast(here("labs", "data", "week8", "landsat_20190412.tif"))
landsat_20190701 <- rast(here("labs", "data", "week8", "landsat_20190701.tif"))
```

#### rename layers
Update the layer names to match the spectral bands they correspond to.
```{r include=TRUE}
names(landsat_20180612) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20180815) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20181018) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20181103) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20190122) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20190223) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20190412) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
names(landsat_20190701) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
```

#### create NDVI function
Here we define a function to compute the Normalized Difference Vegetation Index (NDVI). NDVI computes the difference in reflectance in the near infrared and red bands, normalized by their sum. 

```{r include=TRUE}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
```

#### compute NDVI
Here we apply the NDVI function we created to compute NDVI for each image using the `lapp()` function. The `lapp()` function applies a function to each cell using layers as arguments. Therefore, we need to tell `lapp()` which layers (or bands) to pass into the function. The NIR band is the 4th layer and the red band is the 3rd layer in our raster. In this case, because we defined the NIR band as the first argument and the red band as the second argument in our function, we tell `lapp()` to use the 4th layer first and 3rd layer second. 

```{r include=TRUE}
ndvi_20180612 <- lapp(landsat_20180612[[c(4, 3)]], fun = ndvi_fun)
ndvi_20180815 <- lapp(landsat_20180815[[c(4, 3)]], fun = ndvi_fun)
ndvi_20181018 <- lapp(landsat_20181018[[c(4, 3)]], fun = ndvi_fun)
ndvi_20181103 <- lapp(landsat_20181103[[c(4, 3)]], fun = ndvi_fun)
ndvi_20190122 <- lapp(landsat_20190122[[c(4, 3)]], fun = ndvi_fun)
ndvi_20190223 <- lapp(landsat_20190223[[c(4, 3)]], fun = ndvi_fun)
ndvi_20190412 <- lapp(landsat_20190412[[c(4, 3)]], fun = ndvi_fun)
ndvi_20190701 <- lapp(landsat_20190701[[c(4, 3)]], fun = ndvi_fun)
```

#### combind NDVI layers
Now we can combined the NDVI layers we created into a single raster stack.
```{r include=TRUE}
all_ndvi <- c(ndvi_20180612,
              ndvi_20180815,
              ndvi_20181018,
              ndvi_20181103,
              ndvi_20190122,
              ndvi_20190223,
              ndvi_20190412,
              ndvi_20190701)
```
Let's update the names of each layer to match the date of each image. 
```{r include=TRUE}
names(all_ndvi) <- c("2018-06-12",
                     "2018-08-15",
                     "2018-10-18",
                     "2018-11-03",
                     "2019-01-22",
                     "2019-02-23",
                     "2019-04-12",
                     "2019-07-01")
```

#### read in study sites
```{r include=TRUE}
sites <- st_read(here("labs","data","week8","study_sites.shp"))

tm_shape(ndvi_20180612) +
  tm_raster() +
  tm_shape(sites) +
  tm_polygons()
```

#### extract NDVI at study sites
Here we find the average NDVI within each study site. The output of extract is a data frame with rows that match the study site dataset, so we bind the results to the original dataset. 
```{r include=TRUE}
sites_ndvi <- terra::extract(all_ndvi, sites, fun = "mean")

sites_annotated <- cbind(sites, sites_ndvi)
```

#### clean results
We're done! Except our data is very untidy... Let's tidy it up!\

-   covert to data frame
-   turn from wide to long format
-   turn layer names into date format
```{r include=TRUE}
sites_clean <- sites_annotated %>%
  st_drop_geometry() %>%
  select(-ID) %>%
  pivot_longer(!study_site) %>%
  rename("NDVI" = value) %>%
  mutate("year" = str_sub(name, 2, 5),
         "month" = str_sub(name, 7, 8),
         "day" = str_sub(name, -2, -1)) %>%
  unite("date", 4:6, sep = "-") %>%
  mutate("date" = lubridate::as_date(date))
```

#### plot results
```{r include=TRUE}
ggplot(sites_clean,
       aes(x = date, y = NDVI,
           group = study_site, col = study_site)) +
  geom_line() 
```

