---
title: "EDS 223: week 1 lab"
author: "Ruth Oliver"
output: html_document
date: "2022-10-03"
---
## introduction
In this lab, we'll explore the basics of map-making in R using the **tmap** package.
The following materials are modified from [Chapter 9 of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/adv-map.html)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## prerequisites
Let's install all the necessary packages to reproduce code from Geocomp with R.
```{r}
#install.packages("spDataLarge", repos = "https://geocompr.r-universe.dev")
#install.packages("remotes")
#remotes::install_github("geocompx/geocompkg")
```
Let's load all necessary packages.
```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
remotes::install_github("r-tmap/tmap@v4")
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(ggplot2) # tidyverse data visualization package

```

Let's also read in some data from the **spDataLarge** package to work with later.
```{r include=TRUE}
nz_elev = rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
```

## map-making basics

Let's start with a pre-loaded spatial object representing the states of New Zealand

```{r include=TRUE}
nz
```

-   we're going to start by using the **tmap** package to make some basic maps
-   **tmap** can work with spatial objects of a variety of clasess, meaning it's highly versatile
-   this approach relies on a series of functions that typically start with "tm\_"

the first element is always "tm_shape"

```{r include=TRUE}

```

now let's plot just the boundaries

```{r include=TRUE}

```

and the shapes and boundaries together

```{r include=TRUE}

```

## map objects

-   **tmap** can store maps as *objects*
-   this means that we store a base map and add additional layers later

```{r include = TRUE}

```

-   we can add new shapes on top of the base map
-   when we add a new shape, all subsequent aesthetic functions refer to it, until we add a new shape

In this case, we're adding a layer with information on elevation and this layer to have 70% transparency.

```{r include = TRUE}

```

we can add points designating high points in the country

```{r include = TRUE}

```

## aesthetic basics

-   up until now, we've been working with the default aesthetics
-   there are 2 types of aesthetics: fixed and those that change with the value of a variable
-   **tmap** works differently than **ggplot2** and doesn't use the "aes()" function

Let's start by changing some fixed aesthetics...First, let's change the color used to fill the NZ shapes.

```{r include=TRUE}

```

now change the color of the boundaries

```{r include=TRUE}

```

and the width of the boundary lines

```{r include=TRUE}

```

and the line type of the boundary lines

```{r include=TRUE}

```

all together now!

```{r include=TRUE}

```

Now let's change the colors based on a value. We noticed that the New Zealand dataset has a column with each state's land area

```{r include=TRUE}
nz
```

Let's try to plot the Land_area column. We might think that the following works, but it doesn't!

```{r include=TRUE}

```

Instead, **tmap** is expecting a character string naming the attribute associated with the geometry

```{r include = TRUE}

```

We can also add an argument that updates the title of the legend

```{r include=TRUE}

```

We can even make it more precise using the "expression" function

```{r include=TRUE}

```

## color settings

-   how we set and define colors can radically change the interpretation of our maps
-   the style argument has several options for breaking data into bins
    -   **style = "pretty"** (default) rounds breaks into evenly spaced whole numbers, where possible
    -   **style = "equal"** divides input values into bins with equal range (best for uniform distributions)
    -   **style = "quantile"** puts the same number of observations into each bin
    -   **style = "jenks"** identifies groups of similar values and maximizes differences between bins

Notice how the following maps display the same data, but look quite different    
```{r include=TRUE}

```

```{r include=TRUE}

```

```{r include=TRUE}

```

```{r include=TRUE}

```

We can also define custom bins
```{r include=TRUE}

```

-   in some cases we might not want to use bins
    -   **style = "cont"** displays colors as a continuous spectrum
    -   **style = "cat"** uses a unique vale for each categorical value
```{r include=TRUE}

```

```{r include=TRUE}

```   
  
## map layout
-   now that we have the basics, we can turn to all the other elements that make a cohesive map
-   **tmap** has lots of options, but we will explore just a few

To clearly give readers the context of our map, we can include a compass and scale bar
```{r include=TRUE}

```

Instead of using a compass and scale bar, we could add latitude/longitudes graticules
```{r include=TRUE}

```

We can also update the background color
```{r include=TRUE}

```

## faceted and animated maps
-   we might have data that varies over time and we want to look at the how it changes
-   one approach is by using faceted plots, a series of plots

```{r include=TRUE}

```


## interactive maps
-   **tmap** is especially powerful because it allows us to make interactive maps using the same syntax
-   all we need to do is enter the interactive mode

```{r include=TRUE}

```

To go back to regular plotting, we just need enter plotting mode
```{r include=TRUE}

```

