---
title: 'EDS 223: week 3 lab notes'
author: "Ruth Oliver"
date: "2022-10-12"
output: pdf_document
---


## introduction
In this lab, we'll explore the basics of spatial operations on vector data in R using the **sf** package.
The following materials are modified from [Chapter e of Geocomputation with R by Rovin Lovelace](https://geocompr.robinlovelace.net/spatial-operations.html)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## prerequisites

```{r install, include=TRUE, eval=FALSE}
rm(list = ls())
library(sf)
library(spData)
library(tmap)
library(tidyverse)
```

## spatial subsetting
Spatial subsetting is the process of converting a spatial object into a new object containing only the features that *relate* in space to another object. This is analogous the attribute subsetting that we covered last week. There are many ways to spatially subset in R, so we will explore a few.

Let's start by going back to the New Zealand datasets and find all the high points in the state of Canterbury.  

The command

```{r include=TRUE, eval=FALSE}
canterbury <- nz %>%
  filter(Name == "Canterbury")

# subsets nz_heights to just the features that intersect Canterbury
c_height <- nz_height[canterbury, ]

tm_shape(nz) +
  tm_polygons() +
  tm_shape(nz_height) +
  tm_dots(col = "red")

tm_shape(nz) +
  tm_polygons() +
  tm_shape(canterbury) +
  tm_polygons(col = "blue") +
  tm_shape(c_height) +
  tm_dots(col = "red")

```

The default is to subset to features that intersect, but we can use other operations, including finding features that do not intersect.

```{r include=TRUE, eval=FALSE}
outside_height <- nz_height[canterbury, , op = st_disjoint]

tm_shape(nz) +
  tm_polygons() +
  tm_shape(canterbury) +
  tm_polygons(col = "blue") +
  tm_shape(outside_height) +
  tm_dots(col = "red")
```

We can perform the same operations using topological operators. These operators return matrices testing the relationship between features. 

```{r include=TRUE, eval=FALSE}
sel_sgbp <- st_intersects(x = nz_height, y = canterbury)
sel_sgbp
sel_logical <- lengths(sel_sgbp) > 0

c_height2 <- nz_height[sel_logical, ]

tm_shape(nz) +
  tm_polygons() +
  tm_shape(canterbury) +
  tm_polygons(col = "blue") +
  tm_shape(c_height2) +
  tm_dots(col = "red")
```

We can also use the **st_filter** function in *sf*

```{r include=TRUE, eval=FALSE}
c_height3 <- nz_height %>%
  st_filter(y = canterbury, .predicate = st_intersects)

tm_shape(nz) +
  tm_polygons() +
  tm_shape(canterbury) +
  tm_polygons(col = "blue") +
  tm_shape(c_height3) +
  tm_dots(col = "red")

```

We can change the predicate option to test subset to features that don't intersect

```{r include=TRUE, eval=FALSE}
outside_height2 <- nz_height %>%
  st_filter(y = canterbury, .predicate = st_disjoint)

tm_shape(nz) +
  tm_polygons() +
  tm_shape(canterbury) +
  tm_polygons(col = "blue") +
  tm_shape(outside_height2) +
  tm_dots(col = "red")
```

## spatial joining
Where attribute joining depends on both data sets sharing a 'key' variable, spatial joining uses the same concept but depends on spatial relationships between data sets. 

Let's test this out by creating 50 points randomly distributed across the world and finding out what countries they call in.

```{r include=TRUE, eval=FALSE}
set.seed(2018)
bb <- st_bbox(world)

random_df <- data.frame(
  x = runif(n = 10, min = bb[1], max = bb[3]),
  y = runif(n = 10, min = bb[2], max = bb[4])
)

random_points <- random_df %>%
  st_as_sf(coords = c("x", "y")) %>%
  st_set_crs("EPSG:4326")

tm_shape(world) +
  tm_fill() +
  tm_shape(random_points) +
  tm_dots(col = "red")
```

Let's first use spatial subsetting to find just the countries that contain random points.

```{r include=TRUE, eval=FALSE}
world_random <- world[random_points, ]

tm_shape(world) +
  tm_fill() +
  tm_shape(world_random) +
  tm_fill(col = "red")
```

Now let's perform a spatial join to add the info from each country that a point falls into onto the point dataset.

```{r include=TRUE, eval=FALSE}
random_joined  <- st_join(random_points, world)

tm_shape(world) +
  tm_fill() +
  tm_shape(random_joined) +
  tm_dots(col = "name_long")
```

By default, **st_join** performs a left join. We change this and instead perform an inner join.

```{r include=TRUE, eval=FALSE}
random_joined_inner <- st_join(random_points, world, left = FALSE)
```

## non-overlapping joins
Sometimes we might want join geographic datasets that are strongly related, but do not have overlapping geometries. To demonstrate this, let's look at data on cycle hire points in London.

```{r include=TRUE, eval=FALSE}
tmap_mode("view")

tm_shape(cycle_hire) +
  tm_dots(col = "blue", alpha = 0.5) +
  tm_shape(cycle_hire_osm) +
  tm_dots(col = "red", alpha = 0.5)
```

We can check if any of these points overlap.
```{r include=TRUE, eval=FALSE}
any(st_touches(cycle_hire, cycle_hire_osm, sparse = FALSE))
```

Let's say we need to join the 'capacity' variable in 'cycle_hire_osm' onto the official 'target' data in 'cycle_hire'. The simplest method is using the topological operator **st_is_within_distance()**.

```{r include=TRUE, eval=FALSE}
sel <- st_is_within_distance(cycle_hire, cycle_hire_osm, dist = 20)

summary(lengths(sel) > 0) #summarizes the number of points within 20 meters
```

Now, we'd like to add the values from 'cycle_hire_osm' onto the 'cycle_hire' points.
```{r include=TRUE, eval=FALSE}
z <- st_join(cycle_hire, cycle_hire_osm, st_is_within_distance, dist = 20)

nrow(cycle_hire)
nrow(z)
```
Note: the number of rows of the join is larger than the number of rows in the original dataset. Why? Because some points in 'cycle_hire' were within 20 meters of multiple points in 'cycle_hire_osm'. If we wanted to aggregate so we have just one value per original point, we can use the aggregation methods from last week.

```{r include=TRUE, eval=FALSE}
z <- z %>%
  group_by(id) %>%
  summarise(capacity = mean(capacity))
```

## spatial aggregation
Similar to attribute data aggregation, spatial data aggregation condenses data (we end up with fewer rows than we started with).

Let's say we wanted to find the average height of high point in each region of New Zealand. We could use the **aggregate** function in base R.
```{r include=TRUE, eval=FALSE}
nz_agg = aggregate(x = nz_height, by = nz, FUN = mean)
```
The result of this is an object with the same geometries as the aggregating feature data set (in this case 'nz').

```{r include=TRUE, eval=FALSE}

```

We could also use a **sf**/**dplyr** approach.
```{r include=TRUE, eval=FALSE}
nz_agg <- st_join(nz, nz_height) %>%
  group_by(Name) %>%
  summarise(elevation = mean(elevation, na.rm = TRUE))
```

## joining incongruent layers
We might want to aggregate data to geometries that are not congruent (i.e. their boundaries don't line up). This causes issues when we think about how to summarize associated values.
```{r include=TRUE, eval=FALSE}
head(incongruent)
head(aggregating_zones)

tm_shape(incongruent) +
  tm_polygons() +
  tm_shape(aggregating_zones) +
  tm_borders(col = "red")
```

The simplest method for dealing with this is using area weighted spatial interpolation which transfers values from the 'incongruent' object to a new column in 'aggregating_zones' in proportion with the area of overlap. 

```{r include=TRUE, eval=FALSE}

iv <- incongruent["value"]
agg_aw <- st_interpolate_aw(iv, aggregating_zones, extensive = TRUE)

tm_shape(agg_aw) +
  tm_fill(col = "value")
```

## distance relationships
While topological relationships are binary (features either intersect or don't), distance relationships are continuous. 

We can use the following to find the distance between the highest point in NZ and the centroid of the Canterbury region.

```{r include=TRUE, eval=FALSE}
nz_highest <- nz_height %>%
  slice_max(n = 1, order_by = elevation)

canterbury_centroid = st_centroid(canterbury)
st_distance(nz_highest, canterbury_centroid)

```
Note: this function returns distances with units (yay!) and as a matrix, meaning we could find the distance between many locations at once.
